Simplified integrated nested Laplace approximation for extended latent Gaussian models

### 2022-09-15 src/docs_paper/paper.Rmd
Small fixes to paper

# Background

## Integrated nested Laplace approximation \label{sec:inla}

@@

## Template Model Builder \label{sec:tmb}

@@

The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad].
The "AD" in ADMB is automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
AD is popular in machine learning [@baydin2017automatic], for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
`TMB` uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.

Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb] use the notation $u$ for random effects and $\theta$ for parameters. We aim for consistency with Section \ref{sec:inla}.}
Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
In `TMB`, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
A standard maximum likelihood approach is to optimise
\begin{equation}
L_\ell(\btheta) \triangleq \int_{\mathbb{R}^n} p(\y \, | \, \x, \btheta) \text{d}\x = \int_{\mathbb{R}^n} \exp(-\ell(\x, \btheta)) \text{d}\x \label{eq:flikelihood}
\end{equation}
with respect to $\btheta$ to find the maximum likelihood estimator (MLE) $\hat \btheta$.
Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
\begin{equation}
f(\x, \btheta) 
\triangleq - \log p(\y \, | \, \x, \btheta) p(\x, \btheta)
= \ell(\x, \btheta) - \log p(\x, \btheta),
\end{equation}
equivalent up to an additive constant to the negative log-posterior.
<!-- \begin{equation} -->
<!-- f(\x, \btheta)  -->
<!-- = - \log p(\y, \x, \btheta)  -->
<!-- = - \log p(\x, \btheta \, | \, \y) - C, -->
<!-- \end{equation} -->
<!-- where $C = \log p(\y)$ is the log evidence. -->
Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\btheta$
\begin{equation}
L_f(\btheta) 
\triangleq \int_{\mathbb{R}^n} \exp(-f(\x, \btheta)) \text{d}\x 
\propto \int_{\mathbb{R}^n} p(\x, \btheta \, | \, \y) \text{d}\x = p(\btheta \, | \, \y). \label{eq:1}
\end{equation}
Integrating out the random effects directly, as in Equation \ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section \ref{sec:approxi}.
\begin{equation*}
f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta) = - \frac{\partial^2}{\partial \x^2} \log p(\y, \x, \btheta) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= - \frac{\partial^2}{\partial \x^2} \log p(\x \, | \, \btheta, \y) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= \hat{\bm{Q}}(\btheta).
\end{equation*}
Inference proceeds by optimising $L^\star_f(\btheta)$ via minimisation of
\begin{equation}
-\log L^\star_f(\btheta) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\btheta)) + f(\hat{\bm{\mu}}(\btheta), \btheta)  \label{eq:nllaplace},
\end{equation}
where $\propto$ is used to mean proportional up to an additive constant.
The parameters of the Gaussian approximation (Equation \ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.

# The Naomi small-area estimation model

### Prevalence component

### ANC component

### ART component

# Results \label{sec:results}

# Conclusions \label{sec:conclusions}

## Supporting information

**Appendix A**: Supporting information

## Funding

@@

## Disclaimer

>>>>

# The Naomi model

## Prevalence component

## ANC component

## ART component

# Inference

## Integrated nested Laplace approximation\label{sec:inla}

@@

@wood2020simplified propose a "simplified" version of the INLA algorithm.

## Template Model Builder\label{sec:tmb}

@@

The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad], where the "AD" stands for automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
AD is popular in machine learning [@baydin2017automatic], including as the basis for backpropagation algorithm, and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
`TMB` uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.

Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.^[@kristensen2016tmb use the notation $u$ for random effects and $\theta$ for parameters. We aim for consistency with Section \ref{sec:inla}.]
Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
In `TMB`, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
A standard maximum likelihood approach is to optimise
\begin{equation}
L_\ell(\btheta) \triangleq \int_{\mathbb{R}^n} p(\y \, | \, \x, \btheta) \text{d}\x = \int_{\mathbb{R}^n} \exp(-\ell(\x, \btheta)) \text{d}\x \label{eq:flikelihood}
\end{equation}
with respect to $\btheta$ to find the maximum likelihood estimator (MLE) $\hat \btheta$.
Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
\begin{equation}
f(\x, \btheta) 
\triangleq - \log p(\y \, | \, \x, \btheta) p(\x, \btheta)
= \ell(\x, \btheta) - \log p(\x, \btheta),
\end{equation}
equivalent up to an additive constant to the negative log-posterior.
<!-- \begin{equation} -->
<!-- f(\x, \btheta)  -->
<!-- = - \log p(\y, \x, \btheta)  -->
<!-- = - \log p(\x, \btheta \, | \, \y) - C, -->
<!-- \end{equation} -->
<!-- where $C = \log p(\y)$ is the log evidence. -->
Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\btheta$
\begin{equation}
L_f(\btheta) 
\triangleq \int_{\mathbb{R}^n} \exp(-f(\x, \btheta)) \text{d}\x 
\propto \int_{\mathbb{R}^n} p(\x, \btheta \, | \, \y) \text{d}\x = p(\btheta \, | \, \y). \label{eq:1}
\end{equation}
Integrating out the random effects directly, as in Equation \ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section \ref{sec:approxi}.
\begin{equation*}
f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta) = - \frac{\partial^2}{\partial \x^2} \log p(\y, \x, \btheta) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= - \frac{\partial^2}{\partial \x^2} \log p(\x \, | \, \btheta, \y) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= \hat{\bm{Q}}(\btheta).
\end{equation*}
Inference proceeds by optimising $L^\star_f(\btheta)$ via minimisation of
\begin{equation}
-\log L^\star_f(\btheta) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\btheta)) + f(\hat{\bm{\mu}}(\btheta), \btheta)  \label{eq:nllaplace},
\end{equation}
where $\propto$ is used to mean proportional up to an additive constant.
The parameters of the Gaussian approximation (Equation \ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.

# Case-study\label{sec:results}

# Discussion\label{sec:conclusions}

# Appendix {#appendix .unnumbered}

# Acknowledgements {#acknowledgements .unnumbered}

@@
